---
title: 'Control chart example:  Per cent Daily Positive Tests'
author: "Kevin Little,Ph.D., Informing Ecological Design, LLC"
date: "5/1/2020"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

### Gating Criteria from State DHS
Symptoms 	Downward trajectory of influenza-like illnesses (ILI) reported within a 14-day period. 	Red indicator = no progress

Symptoms 	Downward trajectory of COVID-like syndromic cases reported within a 14-day period. 	Red indicator = no progress

Cases 	Downward trajectory of positive tests as a percent of total tests within a 14-day period. 	Red indicator = no progress

Hospitals 	Criteria are currently under development in partnership with Wisconsin healthcare stakeholders.


### Testing Data:  Plots of Daily Counts
```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(GGally)
library(gridExtra)

knitr::opts_chunk$set(echo = FALSE)
#figure out how to automate the load from the state site https://data.dhsgis.wi.gov/datasets/covid-19-historical-data-table/data
#file manually downloaded on 4-30 at 9 pm.
df1 <- read_csv("COVID-19_Historical_Data_Table_2020_04_30_21_46.csv")


#pull state and county records
df1 <- df1 %>% filter(GEO %in% c('State','County'))

#time records trashed on conversion to CSV
df1$Date_Time <- df1$LoadDttm/1000

class(df1$Date_Time) <- c('POSIXt','POSIXct')
df1$Date_reported <- as.Date(df1$Date_Time)

#distinguish baseline from most recent 14 days
cut_date <- max(df1$Date_reported) - 14

df1$phase <- ifelse(df1$Date_reported <= cut_date,"baseline","last 14 days")

#subset data to look at state-wide numbers
#apparent error in NEGATIVE results on 29 and 30 March
df1_small_state <- df1 %>% select(GEO,NAME,Date_reported,NEGATIVE,POSITIVE,phase) %>% filter(GEO == 'State')
df1_small_state$NEGATIVE[df1_small_state$Date_reported == as.Date("2020-03-29")] <- 15856
df1_small_state$NEGATIVE[df1_small_state$Date_reported == as.Date("2020-03-30")] <- 16550

#now convert cumulative counts to daily counts
df1_small_state <- df1_small_state %>%
                  group_by(NAME) %>% 
  mutate(lag_POSITIVE=lag(POSITIVE)) %>% mutate(POSITIVE_daily = POSITIVE-lag_POSITIVE) %>%
  mutate(lag_NEGATIVE=lag(NEGATIVE)) %>% mutate(NEGATIVE_daily = NEGATIVE - lag_NEGATIVE) %>%
  mutate(Total_daily_tests = POSITIVE_daily + NEGATIVE_daily) %>%
  mutate(POS_pct_daily = 100*(POSITIVE_daily/Total_daily_tests)) 

#knitr::kable(head(df1_small_state))
```





```{r initial_plots, echo=FALSE, warning=FALSE, message=FALSE, fig.height=8}
df1_small_state_longer <- df1_small_state %>% pivot_longer(cols = -c("GEO","NAME","Date_reported","phase"),
                                               names_to = "Measure",
                                               values_to = "value")

dfA <- df1_small_state_longer %>% filter(Measure %in% c('NEGATIVE_daily','POS_pct_daily','POSITIVE_daily','Total_daily_tests'))
dfA$Measure <- factor(dfA$Measure, levels=c("POSITIVE_daily", 'NEGATIVE_daily','Total_daily_tests','POS_pct_daily'))

#create medians
med_A <- as.vector(tapply(dfA$value,dfA$Measure, median, na.rm=TRUE))
Measure <- levels(dfA$Measure)
df.hlines <- data.frame(Measure, med_A)

p1 <- ggplot(data=dfA,aes(x=Date_reported,y=value))+
        theme_bw()+
        geom_point(aes(shape=phase))+
        facet_wrap(~Measure,
                   ncol = 1,
                   scales = 'free_y')+
        labs(title="Test counts",
             subtitle="Dashed lines are medians for entire series")+
        geom_hline(aes(yintercept=med_A),data=df.hlines,lty=2)+
        ylab("")+
        xlab("Date Reported")+
        annotate("rect", fill = "blue", alpha = 0.1, 
            xmin = cut_date + .5, xmax = max(dfA$Date_reported)+.5,
            ymin = -Inf, ymax = Inf)

p1 

```

### Control Chart view:  Daily Percent Positive Tests 
Compute control limits shown via three options:

1. specify all variation as within day (p-chart structure, binomial variation).

2. specify all variation as between days (individuals chart structure, point to point variation using standard control chart scaling)

3. allow variation as combination of within and between day variation per D.B. Laney ((2002), “Improved Control Charts for Attributes,” *Quality Engineering*, **14**(4), 531–537).

As the number of daily tests will likely continue to increase, it makes sense to account for variation in count size.  My current vote is option 3.

In any approach, calculate the control chart center line (average) and the limits using a baseline data series 14 records shorter than the entire record.

Extend the center line and limits based on the baseline calculations.  Examine the chart for signals of special causes of variation.

Plot the 5% gating level to examine relationship of current positive rate to the 5% level.

If there are no special cause signals, on the next day, extend the baseline by 1 day. 

Repeat.



```{r control_charts, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4}
df0 <- df1_small_state %>% filter(Date_reported >= as.Date("2020-03-25"))

df0$pi <- df0$POS_pct_daily/100

df0_baseline <- df0 %>% filter(Date_reported <= max(Date_reported) - 14)

#compute binomial limits
df0$qi <- 1 - df0$pi
p_bar <-sum(df0_baseline$POSITIVE_daily)/sum(df0_baseline$Total_daily_tests)
df0$sigma_pi <- sqrt(p_bar*(1-p_bar)/df0$Total_daily_tests)
df0$p_UCL <- p_bar + 3*df0$sigma_pi
df0$p_LCL <- p_bar - 3*df0$sigma_pi

#compute z scores and assess variation in z scores
df0$zi <- (df0$pi - p_bar)/df0$sigma_pi
Rzi <- abs(diff(df0$zi[df0$Date_reported <= max(df0$Date_reported)-14]))
sigma_zi <- mean(Rzi)/1.128
df0$Laney_sigma <- df0$sigma_pi*sigma_zi
df0$pprime_LCL <- p_bar - 3*df0$Laney_sigma
df0$pprime_UCL <- p_bar + 3*df0$Laney_sigma

#compute I chart values
p_barI <- mean(df0_baseline$pi)  #assumes equal weighting of the Lot proportions
Rpi <- abs(diff(df0_baseline$pi))
sigma_aver_Rpi <- mean(Rpi)/1.128
sigma_median_Rpi <- median(Rpi)/0.9554

pchart <- ggplot(data=df0,aes(x=Date_reported,y=100*pi)) +
  theme_bw()+
  geom_point(aes(shape=phase),size=rel(2))+
  geom_line()+
  labs(title="p control chart for Pct Positive tests",
       x=" ",
       y="%",
       subtitle="Limits based on binomial variation, scaled by daily n; baseline period unshaded") +
  geom_hline(yintercept=100*p_bar)+
  geom_line(aes(x=Date_reported,y=100*p_LCL),linetype='dashed')+
  geom_line(aes(x=Date_reported,y=100*p_UCL),linetype='dashed')+
  geom_hline(yintercept= 5, colour="blue",lty="dotted")+
  ylim(0,20)+
  annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = cut_date + .5, xmax = max(df0$Date_reported)+.5,
        ymin = -Inf, ymax = Inf)+
  scale_shape_discrete(na.translate=FALSE)+
         theme(legend.position = c(0.99, 0.99),
               legend.justification = c("right", "top"),
               legend.text = element_text(size = 6),
               legend.title= element_text(size = 8))

pchart

#ichart
ichart <-ggplot(data=df0,aes(x=Date_reported,y=100*pi)) +
  theme_bw()+
  geom_point(aes(shape=phase),size=rel(2))+
  geom_line()+
  labs(title="Individuals control chart for Pct Positive tests",
       x="",
       y="%",
       subtitle="Limits based between day variation ignoring test counts; baseline period unshaded") +
  geom_hline(yintercept=100*p_barI)+
  geom_hline(yintercept= 5, colour="blue",lty="dotted")+
  ylim(0,20)+
  annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = cut_date + .5, xmax = max(df0$Date_reported)+.5,
        ymin = -Inf, ymax = Inf)+
  scale_shape_discrete(na.translate=FALSE)+
         theme(legend.position = c(0.99, 0.99),
              legend.justification = c("right", "top"),
              legend.text = element_text(size = 6),
              legend.title= element_text(size = 8))

ichart_Median <- ichart+ geom_line(aes(x=Date_reported,y=100*(p_barI - 3*sigma_median_Rpi)),linetype='dashed')+
  geom_line(aes(x=Date_reported,y=100*(p_barI + 3*sigma_median_Rpi)),linetype='dashed')+
  labs(caption="Limits based on median moving range")
#ichart_Median

ichart_Mean <- ichart+ geom_line(aes(x=Date_reported,y=100*(p_barI - 3*sigma_aver_Rpi)),linetype='dashed')+
  geom_line(aes(x=Date_reported,y=100*(p_barI + 3*sigma_aver_Rpi)),linetype='dashed') +
  labs(caption="Limits use average moving range; option is to use median moving range")
ichart_Mean

#make p prime chart
pprime_chart <-ggplot(data=df0,aes(x=Date_reported,y=100*pi)) +
  theme_bw()+
  geom_point(aes(shape=phase),size=rel(2))+
  geom_line()+
  labs(title="p' control chart for Pct Positive",
       x="",
       y="%",
       subtitle="Limits based on Laney calculations; baseline period unshaded") +
  geom_hline(yintercept=100*p_bar)+
  geom_line(aes(x=Date_reported,y=100*pprime_LCL),linetype='dashed')+
  geom_line(aes(x=Date_reported,y=100*pprime_UCL),linetype='dashed')+
  geom_hline(yintercept= 5, colour="blue",lty="dotted")+
  ylim(0,20)+
  scale_shape_discrete(na.translate=FALSE)+
         theme(legend.position = c(0.99, 0.99),
              legend.justification = c("right", "top"),
              legend.text = element_text(size = 6),
              legend.title= element_text(size = 8))+
  annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = cut_date + .5, xmax = max(df0$Date_reported)+.5,
        ymin = -Inf, ymax = Inf)
                
pprime_chart

#number of tests per day
ptests <- ggplot(data=df0,aes(x=Date_reported,y=Total_daily_tests))+
          theme_bw()+
          geom_point(aes(shape=phase),size=rel(2))+
          geom_line()+
          labs(title="Total Daily Tests",
            subtitle="Baseline Median Daily Tests, dashed line",
            x="",
            y="")+
      geom_hline(yintercept=median(df0$Total_daily_tests[df0$Date_reported <= cut_date]),lty="dashed")+
      annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = cut_date + .5, xmax = max(df0$Date_reported)+.5,
        ymin = -Inf, ymax = Inf)+
      scale_shape_discrete(na.translate=FALSE)+
         theme(legend.position = c(0.05, 0.99),
              legend.justification = c("left", "top"),
              legend.text = element_text(size = 6),
              legend.title= element_text(size = 8))
 
ptests



```

### Note on independence assumption

If there are clusters of positive cases violating the independence assumptions for within day variation, this will tend to lead to 'Criterion 1' signal on the control chart; the effective number of observations is reduced in the presence of clusters, which implies that the limits based on independence will be narrower than is the case.

